---
title: "Advanced Features in fmriparametric"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Features in fmriparametric}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(fmriparametric)
```

## Introduction

This vignette covers the advanced features available in `fmriparametric` version 3.0, including:

- K-means spatial clustering for heterogeneous data
- Tiered refinement system for challenging voxels
- Parallel processing capabilities
- Enhanced diagnostics and visualization

These features are designed to handle real-world fMRI data with spatial heterogeneity, poor signal-to-noise ratios, and computational demands of whole-brain analysis.

## K-means Spatial Clustering

When HRF characteristics vary across brain regions, global parameter estimates may not adequately capture local variations. The K-means clustering approach identifies spatial clusters with similar HRF properties and refines estimates within each cluster.

### Basic Usage

```{r kmeans-example, eval=FALSE}
# Enable K-means clustering
fit_kmeans <- estimate_parametric_hrf_v3(
  fmri_data = fmri_data,
  event_model = event_model,
  recenter_kmeans_passes = 2,  # Number of K-means iterations
  kmeans_k = 5,                # Number of clusters
  r2_threshold_kmeans = 0.7,   # Include voxels with R² > 0.7 in clustering
  verbose = TRUE
)
```

### How It Works

1. **Voxel Selection**: High-quality voxels (R² > threshold) are selected for clustering
2. **Parameter Transformation**: Parameters are normalized for clustering (e.g., log-transform width)
3. **Clustering**: K-means identifies clusters in parameter space
4. **Refinement**: Each cluster's centroid becomes a new expansion point for Taylor approximation
5. **Update**: Voxels are re-estimated using their cluster's expansion point

### When to Use K-means

K-means clustering is particularly beneficial when:

- Analyzing whole-brain data with diverse regions
- HRF characteristics show clear spatial patterns
- Standard global refinement plateaus at suboptimal solutions
- You have sufficient high-quality voxels for reliable clustering

## Tiered Refinement System

Not all voxels are equally difficult to fit. The tiered refinement system classifies voxels based on fit quality and applies appropriate refinement strategies.

### Refinement Queues

Voxels are classified into three queues:

1. **Easy** (R² > 0.7, low SE): Well-fit voxels requiring no refinement
2. **Moderate** (0.3 < R² < 0.7): Benefit from local re-centering
3. **Hard** (R² < 0.3 or high SE): Require full Gauss-Newton optimization

### Configuration

```{r refinement-example, eval=FALSE}
# Configure refinement thresholds
fit_refined <- estimate_parametric_hrf_v3(
  fmri_data = fmri_data,
  event_model = event_model,
  refinement_opts = list(
    apply_refinement = TRUE,
    r2_threshold_hard = 0.3,      # Voxels below this are "hard"
    r2_threshold_moderate = 0.7,  # Voxels below this are "moderate"
    se_threshold_hard = 0.5,      # High SE indicates "hard" voxel
    se_threshold_moderate = 0.3,  # Medium SE indicates "moderate"
    max_iter_gn = 10              # Maximum Gauss-Newton iterations
  ),
  compute_se = TRUE,  # Required for SE-based classification
  verbose = TRUE
)

# Check refinement results
summary(fit_refined)
```

### Refinement Strategies

#### Local Re-centering (Moderate Voxels)
- Uses each voxel's current estimate as expansion point
- Single Taylor approximation pass
- Updates only if R² improves
- Computationally efficient

#### Gauss-Newton Optimization (Hard Voxels)
- Full nonlinear optimization
- Iterative parameter updates with line search
- Handles cases far from initial estimates
- More computationally intensive but powerful

## Parallel Processing

For large datasets, parallel processing can significantly reduce computation time.

### Setup

```{r parallel-example, eval=FALSE}
# Automatic core detection
fit_parallel <- estimate_parametric_hrf_v3(
  fmri_data = fmri_data,
  event_model = event_model,
  n_cores = NULL,  # Auto-detect available cores
  progress = TRUE
)

# Specify core count
fit_parallel <- estimate_parametric_hrf_v3(
  fmri_data = fmri_data,
  event_model = event_model,
  n_cores = 8,
  progress = TRUE
)
```

### Parallel Operations

The following operations are parallelized:

1. K-means cluster processing
2. Local re-centering for moderate voxels
3. Gauss-Newton optimization for hard voxels
4. Standard error calculations

### Backend Support

The package automatically selects the best available backend:

- **Unix/macOS**: `future::multicore` or `parallel::mclapply`
- **Windows**: `future::multisession` or `parallel::parLapply`
- **Fallback**: Sequential processing if no parallel backend available

## Rock-Solid Mode

For production environments or challenging datasets, use the bulletproof `estimate_parametric_hrf_rock_solid` function:

```{r rock-solid-example, eval=FALSE}
# Maximum safety mode
fit_safe <- estimate_parametric_hrf_rock_solid(
  fmri_data = fmri_data,
  event_model = event_model,
  safety_mode = "maximum",  # Most conservative
  error_report = TRUE,      # Detailed diagnostics
  verbose = TRUE
)

# Check error recovery
if (fit_safe$metadata$errors_recovered > 0) {
  print(fit_safe$error_report)
}
```

### Safety Modes

- **Maximum**: All safety checks, multiple fallbacks, detailed diagnostics
- **Balanced**: Standard safety features, good performance
- **Performance**: Minimal overhead, for clean data

## Enhanced Diagnostics

### Visualization

The enhanced plot method provides multiple diagnostic views:

```{r plot-example, eval=FALSE}
# View fitted HRF curves
plot(fit_refined, type = "hrf", n_curves = 20)

# Parameter distributions
plot(fit_refined, type = "parameters")

# Diagnostic plots (R² distribution, parameter vs R²)
plot(fit_refined, type = "diagnostic")

# Refinement queue visualization
plot(fit_refined, type = "refinement")
```

### Summary Statistics

The summary method now includes refinement information:

```{r summary-example, eval=FALSE}
summ <- summary(fit_refined)

# Refinement summary
print(summ$refinement_summary)

# R-squared breakdown
cat("Excellent fits (R² > 0.7):", 
    round(100 * summ$r_squared_summary$prop_excellent, 1), "%\n")
cat("Failed fits (R² < 0.1):", 
    summ$r_squared_summary$n_failed, "voxels\n")
```

## Performance Optimization Tips

### 1. Choosing Parameters

```{r optimization-tips, eval=FALSE}
# For speed (large datasets, initial exploration)
fit_fast <- estimate_parametric_hrf_v3(
  fmri_data = fmri_data,
  event_model = event_model,
  recenter_global_passes = 1,    # Fewer global iterations
  recenter_kmeans_passes = 0,    # Skip K-means
  refinement_opts = list(apply_refinement = FALSE),  # No refinement
  compute_se = FALSE,            # Skip SE calculation
  n_cores = parallel::detectCores() - 1
)

# For accuracy (final analysis, publication)
fit_accurate <- estimate_parametric_hrf_v3(
  fmri_data = fmri_data,
  event_model = event_model,
  theta_seed = "data_median",    # Data-driven initialization
  recenter_global_passes = 3,    # More global iterations
  recenter_kmeans_passes = 2,    # K-means clustering
  kmeans_k = 7,                  # More clusters
  refinement_opts = list(
    apply_refinement = TRUE,
    r2_threshold_hard = 0.4,     # More aggressive refinement
    r2_threshold_moderate = 0.8,
    max_iter_gn = 15
  ),
  compute_se = TRUE,
  n_cores = parallel::detectCores() - 1
)
```

### 2. Memory Management

For very large datasets:

```{r memory-tips, eval=FALSE}
# Process ROI separately
roi_mask <- create_mask(roi_coordinates)
fit_roi <- estimate_parametric_hrf_v3(
  fmri_data = fmri_data,
  event_model = event_model,
  mask = roi_mask,  # Reduces memory usage
  n_cores = 4       # Limit cores to control memory
)

# Or use rock-solid mode with automatic chunking
fit_large <- estimate_parametric_hrf_rock_solid(
  fmri_data = fmri_data,
  event_model = event_model,
  safety_mode = "balanced"  # Includes memory management
)
```

### 3. Troubleshooting Poor Fits

If many voxels have poor fits:

```{r troubleshooting, eval=FALSE}
# 1. Check data quality
summ <- summary(fit_initial)
hist(summ$r_squared_summary$summary, main = "R² Distribution")

# 2. Try data-driven initialization
fit_data_init <- estimate_parametric_hrf_v3(
  fmri_data = fmri_data,
  event_model = event_model,
  theta_seed = "data_median"
)

# 3. Adjust parameter bounds
fit_wider_bounds <- estimate_parametric_hrf_v3(
  fmri_data = fmri_data,
  event_model = event_model,
  theta_bounds = list(
    lower = c(1, 0.5, 0),    # Wider bounds
    upper = c(15, 6, 1)
  )
)

# 4. Increase regularization for noisy data
fit_regularized <- estimate_parametric_hrf_v3(
  fmri_data = fmri_data,
  event_model = event_model,
  lambda_ridge = 0.1  # Higher regularization
)
```

## Complete Example

Here's a complete workflow using all advanced features:

```{r complete-example, eval=FALSE}
# Load data
library(fmriparametric)
library(fmrireg)

# Prepare data
fmri_data <- read_fmri("subject01_bold.nii.gz")
events <- read.csv("subject01_events.csv")
event_model <- event_model(onset ~ condition, data = events)

# Initial fit with basic settings
fit_initial <- estimate_parametric_hrf_v3(
  fmri_data = fmri_data,
  event_model = event_model,
  verbose = TRUE
)

# Check initial quality
summary(fit_initial)
plot(fit_initial, type = "diagnostic")

# If needed, apply advanced features
if (mean(fit_initial$r_squared < 0.7) > 0.3) {  # >30% poor fits
  message("Applying advanced refinement...")
  
  fit_advanced <- estimate_parametric_hrf_v3(
    fmri_data = fmri_data,
    event_model = event_model,
    theta_seed = "data_median",
    recenter_global_passes = 3,
    recenter_kmeans_passes = 2,
    kmeans_k = 5,
    refinement_opts = list(
      apply_refinement = TRUE,
      r2_threshold_hard = 0.3,
      r2_threshold_moderate = 0.7,
      max_iter_gn = 10
    ),
    compute_se = TRUE,
    n_cores = parallel::detectCores() - 1,
    verbose = TRUE
  )
  
  # Compare results
  cat("Mean R² improved from", round(mean(fit_initial$r_squared), 3),
      "to", round(mean(fit_advanced$r_squared), 3), "\n")
  
  # Visualize improvements
  plot(fit_advanced, type = "refinement")
}

# Extract results
parameters <- coef(fit_advanced)
param_brain <- unmask(parameters[, "tau"], brain_mask)
write_nifti(param_brain, "tau_map.nii.gz")
```

## Conclusion

The advanced features in `fmriparametric` v3.0 provide powerful tools for handling challenging real-world fMRI data:

- **K-means clustering** adapts to spatial heterogeneity
- **Tiered refinement** efficiently handles voxels of varying difficulty
- **Parallel processing** enables whole-brain analysis
- **Rock-solid mode** ensures reliable results even with pathological data

These features work together to provide accurate, efficient, and robust parametric HRF estimation for modern fMRI analyses.