---
title: "Getting started with parametric HRF estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with parametric HRF estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This short tutorial demonstrates the basic workflow for estimating a parametric
hemodynamic response function (HRF) from an fMRI time series. The underlying
algorithm implements the goals outlined in the project proposal: efficient
voxel-wise estimation of interpretable HRF parameters with optional refinement.

## Simulated example

```{r}
library(fmriparametric)

# Create a small synthetic dataset
set.seed(1)
fmri_data <- matrix(rnorm(40), nrow = 20, ncol = 2)

events <- matrix(0, nrow = 20, ncol = 1)
events[c(5, 15), 1] <- 1

# Estimate HRF parameters
fit <- estimate_parametric_hrf(fmri_data, events, parametric_hrf = "lwu",
                               verbose = FALSE)
summary(fit)
```

## Diagnostic plot

```{r, fig.width=6, fig.height=4}
plot(fit)
```

The estimated HRF shape and parameter distributions help assess model fit. The
package design follows the proposal's aim of providing robust, interpretable
estimates in a few straightforward steps.
