---
title: Getting started with fmriparametric
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with fmriparametric}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r setup, include = FALSE}
if (exists("params") && requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(params$family))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

Standard fMRI analysis assumes a fixed hemodynamic response function (HRF)
across the whole brain. In reality, vascular timing varies by region --- peaks
arrive earlier in some areas, later in others, and undershoot strength differs
too. Ignoring this variability reduces statistical power and can bias
activation maps. `fmriparametric` estimates voxel-specific HRF parameters so
you can characterize this variability directly.

## First example

```{r first-example}
library(fmriparametric)

set.seed(42)
n_time <- 200
n_voxels <- 5

# Simulated BOLD data (time x voxels) and event timing
bold <- matrix(rnorm(n_time * n_voxels, mean = 100, sd = 10),
               nrow = n_time, ncol = n_voxels)
events <- matrix(0, nrow = n_time, ncol = 1)
events[seq(10, 190, by = 20), ] <- 1

fit <- estimate_parametric_hrf(bold, events, verbose = FALSE)
fit
```

One function call, and you have per-voxel HRF parameter estimates. The rest of
this vignette explains what those numbers mean and how to control the
estimation.

## What gets estimated

The package uses the **Lag-Width-Undershoot (LWU)** model, which describes
the HRF with three interpretable parameters:

| Parameter  | Symbol | Typical range | Meaning                           |
|:-----------|:------:|:-------------:|:----------------------------------|
| Lag        | $\tau$    | 4 -- 8 s      | Time to peak response             |
| Width      | $\sigma$    | 1.5 -- 4 s    | Duration / spread of the response |
| Undershoot | $\rho$    | 0.2 -- 0.6    | Strength of the post-stimulus dip |

The mathematical form is a Gaussian peak minus a delayed, broader Gaussian
undershoot:

$$h(t; \tau, \sigma, \rho) = e^{-\frac{(t-\tau)^2}{2\sigma^2}} - \rho \cdot e^{-\frac{(t-\tau-2\sigma)^2}{2(1.6\sigma)^2}}$$

For the full derivation and Taylor linearization details, see
`vignette("technical-details")`.

## Extracting parameters

```{r extract-params}
hrf_params <- coef(fit)
head(hrf_params)
```

Each row is a voxel; columns are the three LWU parameters plus the amplitude
($\beta_0$).

## Checking fit quality

```{r summary-fit}
summary(fit)
```

The summary shows parameter distributions and R-squared values. R-squared
indicates how well the model explains each voxel's time series --- values above
0.5 generally indicate a good fit. With pure noise (as in this toy example),
R-squared values will be low.

## Visualizing estimated HRFs

```{r plot-fit, fig.width = 7, fig.height = 4}
plot(fit)
```

The plot method shows the estimated HRF curves and parameter distributions
across voxels.

## Multi-seed estimation

A single starting point for the Taylor approximation can miss the true HRF
shape when it's far from the default seed. By default, `fmriparametric`
tries **multiple starting points** spanning the HRF parameter space and keeps
the best fit per voxel. This reduces parameter estimation error by 33--45%
compared to single-seed estimation.

```{r multi-seed}
# Multi-seed is enabled by default (multi_seed = TRUE)
# Disable for faster but less accurate estimation:
fit_single <- estimate_parametric_hrf(bold, events,
                                      multi_seed = FALSE,
                                      verbose = FALSE)
```

You can also supply a custom seed grid --- each row is a
$(\tau, \sigma, \rho)$ starting point:

```{r custom-grid, eval = FALSE}
fit_custom <- estimate_parametric_hrf(
  bold, events,
  seed_grid = rbind(
    c(4, 1.5, 0.2),
    c(6, 2.5, 0.4),
    c(8, 3.5, 0.6)
  )
)
```

## Refinement options

After initial estimation, optional refinement stages improve accuracy for
difficult voxels:

```{r refinement, eval = FALSE}
fit_refined <- estimate_parametric_hrf(
  bold, events,
  global_refinement = TRUE,        # Iterate to improve expansion point
  global_passes = 3,               # Number of iterations
  tiered_refinement = "moderate",  # Refine poorly-fit voxels
  compute_se = TRUE                # Delta-method standard errors
)

# Standard errors when compute_se = TRUE
coef(fit_refined, type = "se")
```

The tiered refinement system classifies voxels by fit quality:

- **Easy** (R-squared > 0.7): Light local recentering
- **Moderate** (0.3 < R-squared < 0.7): Re-expansion from each voxel's current estimate
- **Hard** (R-squared < 0.3): Full Gauss-Newton nonlinear optimization

## Design-first workflow

When fitting the same data with different settings, pre-compute the design
matrices once to avoid redundant work:

```{r design-workflow}
design <- create_parametric_design(bold, events)

fit2 <- estimate_parametric_hrf_from_design(design = design, verbose = FALSE)
summary(fit2)
```

## Using with real fMRI data

With real neuroimaging data from the `fmrireg` ecosystem:

```{r real-data, eval = FALSE}
library(fmrireg)

# Load data
fmri_data <- fmridataset::fmri_dataset(
  scans = "bold.nii.gz",
  mask = "brain_mask.nii.gz",
  TR = 2,
  run_length = 300
)

# Define events
events_df <- data.frame(
  onset = c(10, 30, 50, 70, 90),
  duration = rep(1, 5),
  condition = "stimulus"
)

event_model <- fmrireg::event_model(
  onset ~ condition,
  data = events_df,
  sampling_frame = fmri_data
)

# Estimate
fit <- estimate_parametric_hrf(
  fmri_data = fmri_data,
  event_model = event_model,
  verbose = TRUE
)

# Extract parameter maps
params <- coef(fit)
```

## Next steps

- `vignette("parametric-hrf-estimation")` --- deep walkthrough with simulated
  data, visualizations of true vs. estimated HRFs, and parameter interpretation
  guidance
- `vignette("technical-details")` --- full mathematical derivation of the
  Taylor linearization, ridge regularization, and delta-method standard errors
- `?estimate_parametric_hrf` --- complete parameter reference
