---
title: "Introduction to fmriparametric"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to fmriparametric}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The `fmriparametric` package provides tools for estimating parameters of parametric Hemodynamic Response Function (HRF) models from fMRI data. This vignette introduces the basic concepts and demonstrates the core functionality.

## Installation

```{r eval=FALSE}
# Install from GitHub
devtools::install_github("bbuchsbaum/fmrireg")  # Required dependency
devtools::install_github("bbuchsbaum/fmriparametry")
```

## Basic Concepts

### The Hemodynamic Response Function (HRF)

The HRF describes how neural activity translates into the BOLD signal measured by fMRI. Instead of assuming a fixed HRF shape, parametric models allow the HRF to vary across brain regions by estimating interpretable parameters.

### The Lag-Width-Undershoot (LWU) Model

The LWU model characterizes the HRF using three parameters:

1. **τ (tau)**: Lag or time-to-peak
2. **σ (sigma)**: Width of the response
3. **ρ (rho)**: Undershoot amplitude

The mathematical form is:
$$h(t; \tau, \sigma, \rho) = e^{-\frac{(t-\tau)^2}{2\sigma^2}} - \rho \cdot e^{-\frac{(t-\tau-2\sigma)^2}{2(1.6\sigma)^2}}$$

## Basic Usage

### Simulated Example

```{r eval=FALSE}
library(fmriparametric)
library(fmrireg)

# Simulate fMRI data
set.seed(123)
n_time <- 200
n_voxels <- 100
TR <- 2

# Create event timing
events <- data.frame(
  onset = seq(20, 180, by = 40),
  duration = rep(1, 5),
  trial_type = "stimulus"
)

# Generate synthetic BOLD data
# In practice, you would load real fMRI data
Y <- matrix(rnorm(n_time * n_voxels), nrow = n_time)

# Create event model
event_model <- fmrireg::event_model(
  onset ~ trial_type,
  data = events,
  sampling_rate = 1/TR
)

# Create fMRI dataset
fmri_data <- fmrireg::matrix_dataset(
  Y,
  sampling_rate = 1/TR,
  TR = TR
)

# Estimate parametric HRF
fit <- estimate_parametric_hrf(
  fmri_data = fmri_data,
  event_model = event_model,
  parametric_hrf = "lwu",
  verbose = TRUE
)

# View results
print(fit)
```

### Real Data Example

```{r eval=FALSE}
# Load real fMRI data
fmri_data <- fmrireg::read_fmri("path/to/functional.nii.gz")
mask <- fmrireg::read_fmri("path/to/brain_mask.nii.gz")

# Load event timing
events <- read.csv("path/to/events.csv")

# Create event model
event_model <- fmrireg::event_model(
  onset ~ trial_type,
  data = events,
  sampling_rate = 1/TR
)

# Estimate with mask
fit <- estimate_parametric_hrf(
  fmri_data = fmri_data,
  event_model = event_model,
  parametric_hrf = "lwu",
  mask = mask,
  verbose = TRUE
)
```

## Interpreting Results

### Parameter Estimates

```{r eval=FALSE}
# Extract parameters
params <- coef(fit)

# Examine tau (time-to-peak)
hist(params[, "tau"], main = "Distribution of Time-to-Peak", 
     xlab = "Tau (seconds)")

# Examine sigma (width)
hist(params[, "sigma"], main = "Distribution of Response Width",
     xlab = "Sigma (seconds)")

# Examine rho (undershoot)
hist(params[, "rho"], main = "Distribution of Undershoot Amplitude",
     xlab = "Rho")
```

### Summary Statistics

```{r eval=FALSE}
# Get summary
summ <- summary(fit)
print(summ$parameter_summary)
```

## Advanced Options

### Custom Parameter Bounds

```{r eval=FALSE}
# Restrict parameters to specific ranges
fit_constrained <- estimate_parametric_hrf(
  fmri_data = fmri_data,
  event_model = event_model,
  parametric_hrf = "lwu",
  theta_bounds = list(
    lower = c(tau = 3, sigma = 1, rho = 0),
    upper = c(tau = 9, sigma = 4, rho = 0.6)
  )
)
```

### Custom Starting Values

```{r eval=FALSE}
# Use informed starting values
fit_custom_start <- estimate_parametric_hrf(
  fmri_data = fmri_data,
  event_model = event_model,
  parametric_hrf = "lwu",
  theta_seed = c(tau = 5, sigma = 2, rho = 0.3)
)
```

### Adjusting Regularization

```{r eval=FALSE}
# Increase ridge penalty for more stable estimates
fit_regularized <- estimate_parametric_hrf(
  fmri_data = fmri_data,
  event_model = event_model,
  parametric_hrf = "lwu",
  lambda_ridge = 0.1  # Default is 0.01
)
```

## Troubleshooting

### Common Issues

1. **Poor fits in some voxels**: This can occur with low SNR data. Consider:
   - Increasing `lambda_ridge` for more regularization
   - Using a brain mask to exclude non-brain voxels
   - Checking your event timing

2. **Parameters at bounds**: If many voxels have parameters at the bounds:
   - Review your bound settings
   - Check data quality
   - Consider different starting values

3. **Slow performance**: For large datasets:
   - Use a mask to reduce voxel count
   - Future versions will support parallel processing

## Next Steps

- Explore the parameter maps to identify regions with distinct HRF properties
- Compare HRF parameters between conditions or groups
- Use the estimated HRFs for more accurate GLM analyses

## References

For more details on the Taylor approximation method and parametric HRF models, see:

- [Reference to relevant paper once published]
- fmrireg package documentation: https://github.com/bbuchsbaum/fmrireg