---
title: "Quick Start Guide: Parametric HRF Estimation"
author: "fmriparametric package authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start Guide: Parametric HRF Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Installation

```{r install, eval=FALSE}
# Install from GitHub (requires devtools)
devtools::install_github("bbuchsbaum/fmriparametric")
```

## Basic Usage

The `fmriparametric` package estimates voxel-specific HRF parameters using the Lag-Width-Undershoot (LWU) model.

### Minimal Example

```{r minimal}
library(fmriparametric)

# Simulate simple data
set.seed(123)
n_time <- 100
n_voxels <- 10

# Create BOLD signal matrix (time x voxels)
bold_data <- matrix(rnorm(n_time * n_voxels, mean = 100, sd = 10), 
                    nrow = n_time, ncol = n_voxels)

# Create event timing (1 = event, 0 = no event)
events <- rep(0, n_time)
events[c(10, 30, 50, 70, 90)] <- 1
event_matrix <- matrix(events, ncol = 1)

# Estimate HRF parameters
fit <- estimate_parametric_hrf(
  fmri_data = bold_data,
  event_model = event_matrix,
  parametric_hrf = "lwu"
)

# View results
print(fit)
summary(fit)

# Extract parameters
params <- coef(fit)
head(params)
```

### What You Get

The LWU model estimates three parameters per voxel:

- **tau (τ)**: Time to peak in seconds (normal: 4-8s)
- **sigma (σ)**: Width/duration of response (normal: 1.5-4s)  
- **rho (ρ)**: Undershoot amplitude (normal: 0.2-0.6)

## Using with fmrireg

For real fMRI analysis, integrate with the `fmrireg` package:

```{r fmrireg-example, eval=FALSE}
library(fmrireg)

# Load your fMRI data
fmri_data <- fmri_dataset(
  scans = "path/to/functional.nii.gz",
  mask = "path/to/mask.nii.gz",
  TR = 2,
  run_length = 300
)

# Define your events
events_df <- data.frame(
  onset = c(10, 30, 50, 70, 90),  # in seconds
  duration = rep(1, 5),            # event duration
  condition = "stimulus"
)

# Create event model
event_model <- event_model(
  ~ hrf(condition, basis = "spmg1"),  # placeholder basis
  data = events_df,
  sampling_frame = fmri_data
)

# Estimate parametric HRFs
fit <- estimate_parametric_hrf(
  fmri_data = fmri_data,
  event_model = event_model,
  parametric_hrf = "lwu",
  verbose = TRUE
)
```

## Key Options

### Better Initialization

Use data-driven starting values:

```{r data-driven, eval=FALSE}
fit <- estimate_parametric_hrf(
  fmri_data = bold_data,
  event_model = event_matrix,
  theta_seed = "data_driven"  # Instead of default values
)
```

### Global Refinement

Iterate to improve estimates:

```{r refinement, eval=FALSE}
fit <- estimate_parametric_hrf(
  fmri_data = bold_data,
  event_model = event_matrix,
  global_refinement = TRUE,
  global_passes = 3  # Number of iterations
)
```

### Standard Errors

Get uncertainty estimates:

```{r standard-errors, eval=FALSE}
fit <- estimate_parametric_hrf(
  fmri_data = bold_data,
  event_model = event_matrix,
  compute_se = TRUE
)

# Extract standard errors
se_params <- coef(fit, type = "se")
```

### Parallel Processing

Speed up large analyses:

```{r parallel, eval=FALSE}
fit <- estimate_parametric_hrf(
  fmri_data = bold_data,
  event_model = event_matrix,
  parallel = TRUE,
  n_cores = 4
)
```

## Interpreting Results

### Check Fit Quality

```{r fit-quality, eval=FALSE}
# R-squared values indicate fit quality
r2_values <- fit$fit_quality$r_squared

# Good fits have R² > 0.5
good_voxels <- which(r2_values > 0.5)
percent_good <- 100 * length(good_voxels) / length(r2_values)

cat("Percentage of well-fit voxels:", round(percent_good, 1), "%\n")
```

### Visualize Parameters

```{r visualize, eval=FALSE}
# If you have spatial information
library(neuroim2)

# Create parameter maps
tau_map <- NeuroVol(geometry = mask_geometry, data = params[, "tau"])
sigma_map <- NeuroVol(geometry = mask_geometry, data = params[, "sigma"])
rho_map <- NeuroVol(geometry = mask_geometry, data = params[, "rho"])

# Write to file
write_vol(tau_map, "tau_map.nii.gz")
write_vol(sigma_map, "sigma_map.nii.gz")
write_vol(rho_map, "rho_map.nii.gz")
```

## Tips for Success

1. **Start Simple**: Use default settings first, then add complexity

2. **Check Your Events**: Ensure event timing aligns with your TR

3. **Quality Control**: Always check R² values and parameter distributions

4. **Mask Wisely**: Use a brain mask to avoid fitting noise

5. **Be Patient**: Large datasets can take time; use parallel processing

## Getting Help

- **Documentation**: `?estimate_parametric_hrf`
- **Vignettes**: `browseVignettes("fmriparametric")`
- **Issues**: https://github.com/bbuchsbaum/fmriparametric/issues

## Next Steps

For a deeper understanding of the method, see the full vignette:

```{r next-steps, eval=FALSE}
vignette("parametric-hrf-estimation", package = "fmriparametric")
```
